---
title: 'Add average value'
author: 'Sara Lindersson'
date: '`r Sys.Date()`'
output: github_document
---

This script calculates the average values (timmean) for each grid cell from the original daily maximum (TX) and minimum (TN) temperature time series. It then adds these average values to the detrended TN and TX time series.

```{r libraries, warning=F, message=F, collapse=T}
library(glue)
library(knitr)
```

```{r prevent-run, include=F}
knitr::opts_chunk$set(eval = F)
```

```{r setup, include=T}
# Set working directory
knitr::opts_knit$set(root.dir = normalizePath('F:/mswx/data-processed/')) 
```

### First set of parameters
```{r parameters-1}
# Directories TN
ph_in_tn <- 'from-03/tn_tiles'
ph_out_tn <- 'from-05/tn_tiles_mean'

# Create output directory if it doesn't exist
if (!dir.exists(ph_out_tn)) {
  dir.create(ph_out_tn, recursive = TRUE)
}

# Directories TX
ph_in_tx <- 'from-03/tx_tiles'
ph_out_tx <- 'from-05/tx_tiles_mean'

# Create output directory if it doesn't exist
if (!dir.exists(ph_out_tx)) {
  dir.create(ph_out_tx, recursive = TRUE)
}
```

### Calculate temporal mean for TN
```{r tn-mean, message=F, warning=F}
# List all the .nc files in the input directory
files <- list.files(ph_in_tn, pattern = "\\.nc$", full.names = TRUE)

# Loop through each file and compute the average using CDO timmean
for (file in files) {
  # Get the file name
  file_name <- basename(file)
  
  # Define the output file path
  out_file <- file.path(ph_out_tn, paste0("avg_", file_name))

  # Construct the CDO command to calculate the average (timmean)
  cmd <- paste('wsl cdo -O -z zip_1 -b 32 -timmean', file, out_file)
  
  # Execute the CDO command
  shell(cmd = cmd)
}
```

### Calculate temporal mean for TX
```{r tx-mean, message=F, warning=F}
# List all the .nc files in the input directory
files <- list.files(ph_in_tx, pattern = "\\.nc$", full.names = TRUE)

# Loop through each file and compute the average using CDO timmean
for (file in files) {
  # Get the file name
  file_name <- basename(file)
  
  # Define the output file path
  out_file <- file.path(ph_out_tx, paste0("avg_", file_name))

  # Construct the CDO command to calculate the average (timmean)
  cmd <- paste('wsl cdo -O -z zip_1 -b 32 -timmean', file, out_file)
  
  # Execute the CDO command
  shell(cmd = cmd)
}
```

### Second set of parameters
```{r parameters-2}

# Directories TN
ph_in_detr_tn <- 'from-04/tn_tiles_detr' # Tiles with detrended time series
ph_in_avg_tn <- 'from-05/tn_tiles_mean' # Tiles with timmean
ph_out_tn <- 'from-05/tn_tiles_detr_mean'

# Create output directory if it doesn't exist
if (!dir.exists(ph_out_tn)) {
  dir.create(ph_out_tn, recursive = TRUE)
}

# Directories TX
ph_in_detr_tx <- 'from-04/tx_tiles_detr' # Tiles with detrended time series
ph_in_avg_tx <- 'from-05/tx_tiles_mean' # Tiles with timmean
ph_out_tx <- 'from-05/tx_tiles_detr_mean'

# Create output directory if it doesn't exist
if (!dir.exists(ph_out_tx)) {
  dir.create(ph_out_tx, recursive = TRUE)
}
```

### Add average to detrended for TN
```{r tn-add-mean, message=F, warning=F}
# List all the .nc files in the input directory
files <- list.files(ph_in_detr_tn, pattern = "\\.nc$", full.names = TRUE)

# Loop through each file and compute the average using CDO timmean
for (file in files) {
  # Get the file name (without path)
  file_name <- sub('^detr_', '', basename(file))
  
  avg_file <- file.path(ph_in_avg_tn, paste0('avg_', file_name))
  
  # Define the output file path
  out_file <- file.path(ph_out_tn, paste0("detr_avg_", file_name))

  # Construct the CDO command to add the average to detrended
  cmd <- paste('wsl cdo -O -z zip_1 -b 32 -add', file, avg_file, out_file)
  
  # Execute the CDO command
  shell(cmd = cmd)
}
```

### Add average to detrended for TX
```{r tx-add-mean, message=F, warning=F}
# List all the .nc files in the input directory
files <- list.files(ph_in_detr_tx, pattern = "\\.nc$", full.names = TRUE)

# Loop through each file and compute the average using CDO timmean
for (file in files) {
  # Get the file name (without path)
  file_name <- sub('^detr_', '', basename(file))
  
  avg_file <- file.path(ph_in_avg_tx, paste0('avg_', file_name))
  
  # Define the output file path
  out_file <- file.path(ph_out_tx, paste0("detr_avg_", file_name))

  # Construct the CDO command to add the average to detrended
  cmd <- paste('wsl cdo -O -z zip_1 -b 32 -add', file, avg_file, out_file)
  
  # Execute the CDO command
  shell(cmd = cmd)
}
```
End of script.