---
title: 'Extracting data on population and cell area'
author: 'Sara Lindersson'
date: '`r Sys.Date()`'
output: github_document
---

This script extracts grid point data on yearly population figures cell areas for each _disno_ and _subdivision_. Before this, the script resamples the original population maps to align with the meteorological data, and linearly interpolates the 5-year values to annual ones. The script finnaly joins these estimates with the meteorological daily statistics.

```{r libraries, warning=F, message=F, collapse=T}
library(here)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(stringr)
library(raster)
library(zoo)
library(purrr)
library(dplyr)
```

```{r prevent-run, include=F}
knitr::opts_chunk$set(eval = F) # Change to TRUE if chunks should be run when document is knitted.
```

```{r setup, include=T}
# Execute all notebook chunks in the grandparent folder of the notebook
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

### Parameters
```{r parameters}
# Directory to meteorological data
ph_mswx <- 'F:/mswx/temp/3hourly/'

# Directory to population data
ph_ghs <- 'F:/ghs/'

ph_in <- here('data-processed', 'temporary', '04')
ph_out <- here('data-processed', 'temporary', '05')

# Check if output-directory exists, otherwise create it
if (!file.exists(ph_out)) {
  dir.create(ph_out, recursive = T)
}
```

### Import sample of subdivisions for disaster records
Loads `subnat.rds` from script `01-emdat-gdis-link.Rmd`. This dataframe holds the sample of EM-DAT entries at the subnational level, with one row per _disno_ and administrative _subdivision_, accompanied by simplified polygons.
```{r load-subnat}
subnat <- readRDS(here('data-processed','subnat.rds'))
```

### Import population maps
```{r load-pop}
# Years of population data
pop_yrs <- seq(from = 1975, to = 2020, by = 5)

ghs <- terra::rast(paste0(
    ph_ghs,
    'GHS_POP_E', pop_yrs, '_GLOBE_R2023A_4326_30ss_V1_0',
    '/',
    'GHS_POP_E', pop_yrs, '_GLOBE_R2023A_4326_30ss_V1_0.tif'
  )
)
```

### Resample population maps
The script now resamples the population maps from the Global Human Settlement Layer to align with the grid cells of MSWX (the meteorological data), using _exactextractr_.

```{r resample-pop}
# Derive area per grid cell of MSWX
aa <- terra::cellSize(
  terra::rast(
    # Select the first file in directory
    list.files(ph_mswx, pattern = '\\.nc$', full.names = T)[1]
  )
)

# Resample population maps
ghs_rs <- map(pop_yrs, function(yr){
  pop_raster <- ghs[[paste0('GHS_POP_E', yr, '_GLOBE_R2023A_4326_30ss_V1_0')]]
  resampled <- exactextractr::exact_resample(pop_raster, aa, fun = 'sum')
  names(resampled) <- paste0('GHS_POP_E', yr, '_GLOBE_R2023A_4326_30ss_V1_0')
  resampled
})

# Combine all the resampled rasters into a single list
ghs_rs <- do.call(c, ghs_rs)

# Combine with cell size raster
ghs_rs <- c(ghs_rs, aa)

```

### Derive area- and population-estimates per event and admin unit
The script iterates over the subnational records in _subnat_ to derive corresponding area and yearly population estimates at the grid cell level. The resampled maps produce slightly different yearly population totals for each administrative unit. To address this, the script scales the resampled totals to ensure they align with the original values (per administrative unit) before interpolating the yearly population values.

```{r area-pop-grid}
for (i in 1:nrow(subnat)){
  subnat_i <- subnat[i,]
  disno_i <- subnat_i$disno
  gadm_gid_i <- subnat_i$gadm_gid
  
  # Compare yearly population totals to retrieve yearly scaling ratio
  ghs_i <- exact_extract(ghs, subnat_i$geometry, include_xy = T)[[1]] %>%
    rename_with(~ str_replace(.,
                              'GHS_POP_E(\\d+)_GLOBE_R2023A_4326_30ss_V1_0',
                              'p\\1'
    ), 
    starts_with('GHS_POP_E')) %>%
    rename(cf = coverage_fraction) %>%
    pivot_longer(
      cols = -c(x, y, cf),
      names_to = 'year',
      values_to = 'pop'
    ) %>%
    mutate(
      pop = (pop * cf), # Scale population value with coverage fraction
      year = as.POSIXct(
        paste0(as.integer(str_sub(year, 2)), '-01-01'),
        tz = 'GMT'
      )
    ) %>%
    group_by(year) %>%
    summarise(pop_sum_original = sum(pop, na.rm = T))
  
  ghs_rs_i <- exact_extract(ghs_rs, subnat_i$geometry, include_xy = T)[[1]] %>%
    rename_with(~ str_replace(.,
                              'GHS_POP_E(\\d+)_GLOBE_R2023A_4326_30ss_V1_0',
                              'p\\1'), 
                starts_with('GHS_POP_E')) %>%
    rename(cf = coverage_fraction) %>%
    pivot_longer(
      cols = -c(x, y, cf, area),
      names_to = 'year',
      values_to = 'pop'
    ) %>%
    mutate(
      pop = (pop * cf), # Scale population value with coverage fraction
      year = as.POSIXct(
        paste0(as.integer(str_sub(year, 2)), '-01-01'),
        tz = 'GMT'
      )
    ) %>%
    group_by(year) %>%
    summarise(pop_sum_resampl = sum(pop, na.rm = T))
  
  ratios <- left_join(ghs_i, ghs_rs_i, by = 'year') %>%
    mutate(
      ratio = pop_sum_original / pop_sum_resampl
    )
  
  rm(ghs_i, ghs_rs_i)
  
  # Scale the resampled data and interpolate yearly estimates
  x_i <- exact_extract(ghs_rs, subnat_i$geometry, include_xy = T)[[1]] %>%
    rename_with(~ str_replace(.,
                              'GHS_POP_E(\\d+)_GLOBE_R2023A_4326_30ss_V1_0',
                              'p\\1'), 
                starts_with('GHS_POP_E')) %>%
    rename(cf = coverage_fraction) %>%
    mutate(
      y = round(y, 2),
      x = round(x, 2),
      cell_id = row_number(),
      area_km2 = ((area * cf) * 1e-6), # Scale area with cf
      area = NULL
    ) %>%
    # Create columns for every year
    bind_cols(
      purrr::map_dfc(
        c(1976:1979, 1981:1984, 1986:1989, 1991:1994, 1996:1999, 
          2001:2004, 2006:2009, 2011:2014, 2016:2019),
        ~setNames(data.frame(NA),
                  paste0("p", .))
      )
    ) %>%
    pivot_longer(
      cols = -c(cell_id, x, y, cf, area_km2),
      names_to = 'year',
      values_to = 'pop'
    ) %>%
    mutate(
      pop = (pop * cf), # Scale pop with cf
      year = as.POSIXct(
        paste0(as.integer(str_sub(year, 2)), '-01-01'),
        tz = 'GMT'
      )
    ) %>%
    arrange(cell_id, year) %>%
    # Join yearly scaling ratios
    left_join(
      ratios[c('year', 'ratio')],
      by = 'year'
    ) %>%
    # Scale resampled population numbers
    mutate(pop_s = pop * ratio) %>% 
    group_by(cell_id) %>%
    # Interpolate to yearly values
    mutate(
      pop_s_int = na.approx(pop_s, maxgap = 5, na.rm = FALSE)
    ) %>%
    ungroup() %>%
    mutate(year = year(year)) %>%
    # Convert population to integer
    mutate(
      pop = as.integer(pop_s_int)
    ) %>%
    # Drop temporary columns
    dplyr::select(
      -pop_s_int,
      -ratio,
      -pop_s,
      -cell_id,
      -cf
    )
  
  # Load meteorological daily data and link to population and area
  file_i <- paste0(disno_i,'_',gadm_gid_i,'.rds')
  
  data_i <- readRDS(file = paste0(ph_in, '/', file_i)) %>%
    mutate(
      year = year(date)
    ) %>%
    left_join(
      x_i,
      by = c('x', 'y', 'year')
    ) %>%
    mutate(year = NULL)
  
  # Export R-object
  saveRDS(data_i, file = file.path(ph_out, paste0(disno_i, '_', gadm_gid_i, '.rds')))
}
```
End of script.