---
title: 'Extract Percentile Thresholds and Link to Daily Data'
author: 'Sara Lindersson'
date: '`r Sys.Date()`'
output: github_document
---

This script extracts climatology thresholds for air temperature for each grid cell and day of the year, joining this information with the daily statistics of meteorological and population variables. The thresholds of interest are:  
+ pct10: the 10th percentile of the daily minimum air temperature  
+ pct05: the 5th percentile of the daily minimum air temperature  
+ pct90: the 90th percentile of the daily maximum air temperature  
+ pct95: the 95th percentile of the daily maximum air temperature

The output from this script is one aggregated dataframe per _disno_ and _subdivision_, with daily information of meteorological statistics, population numbers and pct thresholds. The script also removes the temporary files produced from scripts 02-05. 

```{r libraries, warning=F, message=F, collapse=T}
library(here)
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(stringr)
library(glue)
```

```{r prevent-run, include=F}
knitr::opts_chunk$set(eval=F) # Change to TRUE if chunks should be run when document is knitted.
```

```{r setup, include=T}
# Execute all notebook chunks in the grandparent folder of the notebook
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

### Parameters
```{r parameters}
# Percentiles
tn <- c('10', '05') # For cold waves
tx <- c('90', '95') # For heat waves

# File pathways
ph_in <- here('data-processed', 'temporary', '05')
ph_out <- here('data-processed', 'daily-data')

# Check if output-directory exists, otherwise create it
if (!file.exists(ph_out)) {
  dir.create(ph_out, recursive = T)
}

# File pathway to percentiles
ph_pct <- 'F:/mswx/data-processed/06/percentiles'
```

### Import subnational disaster records
Loads the output from script _01-emdat-gdis-link.Rmd_.
```{r load-subnat}
subnat <- readRDS(here(
  'data-processed', 'subnat.rds'
))
```

### Data processing and final export of daily time series
```{r data-processing, warning=F}
setwd(ph_pct)

# Loop over subnational records
for (i in 1:nrow(subnat)){
  subnat_i <- subnat[i,]
  disno_i <- subnat_i$disno
  gadm_gid_i <- subnat_i$gadm_gid
  file_i <- paste0(disno_i,'_',gadm_gid_i,'.rds')
  x_i <- readRDS(file = paste0(ph_in, '/', file_i))
  
  # Adjust day of year if leap year
  x_i <- x_i %>%
    mutate(
      leap = leap_year(date),
      doy = yday(date),
      doy = as.integer(
        ifelse(leap == T & doy >= 60, doy-1, doy)
        )
    )
  
  # Choose corresponding percentiles from disaster type
  if (subnat_i$type == 'Cold wave'){
    pcts <- tn
    stat <- 'tn'
  }
  if (subnat_i$type == 'Heat wave'){
    pcts <- tx
    stat <- 'tx'
  }
  
  # Loop over percentiles
  for (pct in pcts){
    
    # Read raster with percentiles
    r <- terra::rast(
      glue('c{stat}{pct}pct.nc'))
    
    r_dat <- exactextractr::exact_extract(
      r,
      subnat_i$geometry,
      include_xy = T
    )[[1]] %>%
      dplyr::select(
        -coverage_fraction
      ) %>%
      pivot_longer(
        cols = -c(x, y),
        names_to = 'doy',
        values_to = paste0('c', stat, pct)
      ) %>%
      mutate(
        # Remove redundant text
        doy = as.integer(gsub('air_temperature_', '', doy)),
        # Round coordinates
        x = round(x, digits = 2),
        y = round(y, digits = 2)
      )
    
    # Join with daily statistics 
    x_i <- x_i %>%
      left_join(r_dat, by = c('x','y','doy'))
  }
  
  # Drop temporary variables
  x_i <- x_i %>%
    dplyr::select(
      -doy,
      -leap
    )
  
  # Export R object
  saveRDS(x_i, file = file.path(ph_out, paste0(disno_i,'_',gadm_gid_i,'.rds')))
}
```

### Delete temporary files
The script now removes the temporary files generated by the data processing scripts 02-05.
```{r delete-temporary}
# unlink(here('data-processed', 'temporary'), recursive = T)
```

End of script.